---
title: 'Assignment 5: Incarceration'
output: html_document
date: "2024-05-25"
---
```{r}
library(ggplot2)
library(dplyr)
library(readr)
```

### Introduction
In this analysis, I want to examine prison incarceration rates for blacks and whites in 1990 in Washington state. By focusing on these demographic groups, I sought to uncover underlying disparities and patterns of racial inequality in the criminal justice system at the state and national levels. The selected years mark a key moment in history, providing a basis for comparison with more current data to assess how incarceration trends may have changed due to policy changes, social movements, or changes in society's attitudes toward crime and punishment. Understanding these dynamics can help address the systemic racial disparities that persist in the U.S. incarceration system. This comparative approach will help highlight the differences and similarities between the two groups in their impact on judicial practice

### Summary Information
```{r}
data_wa <- read.csv("https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv")

data_wa_1990 <- filter(data_wa, year == 1990)

summary_wa <- data_wa_1990 %>%
  summarise(
    Avg_Black_Rate_WA = mean(black_jail_pop_rate, na.rm = TRUE),
    Max_Black_Rate_WA = max(black_jail_pop_rate, na.rm = TRUE),
    Min_Black_Rate_WA = min(black_jail_pop_rate, na.rm = TRUE),
    Avg_White_Rate_WA = mean(white_jail_pop_rate, na.rm = TRUE),
    Max_White_Rate_Wa = max(white_jail_pop_rate, na.rm = TRUE),
    Min_White_Rate_WA = min(white_jail_pop_rate, na.rm = TRUE)
  )

print(summary_wa)

```
In 1990, Washington state data showed stark differences in incarceration rates for blacks and whites. On average, black incarceration rates are about 1,759 per 100,000 people, much higher than the white incarceration rate of 177 per 100,000 people. Blacks have the highest incarceration rate, an alarmingly high 7,266 per 100,000, while whites have a much lower incarceration rate of 348 per 100,000. Interestingly, the lowest black incarceration rate is zero, indicating that some counties have no blacks incarcerated, while the lowest white incarceration rate is about 19 per 100,000 people. The numbers highlight significant racial disparities in incarceration rates and point to deeper problems in the justice system that disproportionately affect black communities.


### The Dataset
The data was collected by vera, which is powered by hundreds of researchers and advocates working to transform the criminal law and immigration systems. The Vera Institute uses a variety of methods to collect and generate data, including working with government agencies, direct data requests and analysis of existing records. The purpose of collecting this data is to inform policy changes, support advocacy efforts, and advance justice reforms, specifically aimed at reducing excessive crime and the negative impact of the justice system on marginalized communities.There are about 91,031 rows and 23 columns in the dataset that I use. 

### Trends Over Time Chart
```{r}
trends_wa <- data_wa %>%
  group_by(year) %>%
  summarise(
    Avg_Black_Rate = mean(black_prison_pop_rate, na.rm = TRUE),
    Avg_White_Rate = mean(white_prison_pop_rate, na.rm = TRUE)
  )

ggplot(trends_wa, aes(x = year)) +
  geom_line(aes(y = Avg_Black_Rate, color = "Black Population"), size = 1.2) +
  geom_line(aes(y = Avg_White_Rate, color = "White Population"), size = 1.2) +
  scale_color_manual(values = c("Black Population" = "black", "White Population" = "blue")) +
  labs(
    title = "Trends in Prison Incarceration Rates by Race in Washington State",
    x = "Year",
    y = "Average Incarceration Rate per 100,000",
    color = "Population Group"
  ) +
  theme_minimal()

ggsave("trends_in_prison_incarceration_rates_wa.png")

```

### Variable Comparison Chart
```{r}
data_wa_1990 <- data_wa %>%
  filter(year == 1990) %>%
  select(county_name, black_prison_pop_rate, white_prison_pop_rate)

ggplot(data_wa_1990, aes(x = black_prison_pop_rate, y = white_prison_pop_rate)) +
  geom_point(aes(color = county_name), alpha = 0.6, size = 4) +
  geom_smooth(method = "lm", se = FALSE, color = "blue") + 
  scale_color_viridis_d() + 
  labs(title = "Comparison of Black vs. White Prison Incarceration Rates in Washington State (1990)",
       x = "Black Prison Incarceration Rate (per 100,000)",
       y = "White Prison Incarceration Rate (per 100,000)",
       color = "County") +
  theme_minimal() +
  theme(legend.position = "right")

ggsave("black_vs_white_incarceration_rates.png")

```


### Map
```{r}
incarceration_data <- read_csv("https://github.com/melaniewalsh/Neat-Datasets/raw/main/us-prison-jail-rates-1990-WA.csv")

black_rates <- incarceration_data %>%
  select(county_name, black_jail_pop_rate) %>%
  arrange(desc(black_jail_pop_rate))  

ggplot(black_rates, aes(x = reorder(county_name, black_jail_pop_rate), y = black_jail_pop_rate)) +
  geom_col(fill = "steelblue") +
  coord_flip() +  
  labs(
    title = "Black Incarceration Rates by County in Washington State (1990)",
    x = "County",
    y = "Incarceration Rate per 100,000",
    caption = "Data source: Neat Datasets"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust text angle for better legibility

ggsave("black_incarceration_rates_by_county.png", width = 10, height = 8)

```

